$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in the Presence of Non-Functional Properties

### 2014-05-07 JournalToSubmit2013/modelingWithPISODM.tex
final version (V2)

%(Aquí lo dejaría con la estructura que teníamos para Caise, presentamos el framework y posteriormente comentamos modelos y transformaciones)

This section presents \pisodm, an MDD-based methodology for building service compositions with non-functional requirements. 
\pisodm provides meta-models for modelling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}). 
Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
The refinement process is driven by transformation rules specified between the meta-models.

@@

This level focusses on the highest-level view of the system, including its business and requirement specifications.
At this stage of the development, the structure and system processing details are still unknown or undetermined.  
\pisodm uses the \textit{e$^3$value}~\cite{Gordijn02valuebased} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.

@@

\begin{example}[To Publish Music]\label{ex:toPublicMusic}
Let us consider the scenario ``To Publish Music'', which will be used as a running example in this section:
An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts. 
In this way, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
The following (external) services will be used by the application:
\begin{itemize}
\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item Facebook and Twitter services export methods for  updating the status of a given user:
\begin{itemize} 
%\item{\sf\small get-Status ( usedid ): String ; 
\item {\sf\small update-Status ( userid, new-status ): String}; 
\end{itemize}
\end{itemize}

@@

The CIM level models are the basis for the development of the application. 
The information presented at this level is (manually) refined into PIM-level models of the methodology \pisodm described next.
Notice that, at this level, the models are (very informally) used to describe both functional and non-functional requirements.

@@

This level focusses on the system functionality, hiding the details of any particular platform.
The specification defines those parts of the system that do not change from one platform to another. 
Our methodology defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess} and \textit{$\pi$-ServiceComposition}.

@@

Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-functional requirements.
The notion of \textit{policy} is used to describe NFRs. 
(In Figure~\ref{fig:CIM:usecasemetamodel}, we have highlighted the concepts related to NFRs).
The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing non-functional requirements with  the concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc non-functional requirement}, {\sc non-functional attribute} and {\sc Constraint}. An {\sc End Consumer} is represented by an {\sc Actor}. An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
 The {\sc Business Collaborator} concept is represented  through {\sc Packages}. 
A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modelled. 
Each {\sc Business Collaborator} combines the features described in each  {\sc Package}. 
Thus, the services functions can be specified being grouped into packages.

@@

 The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented as part of description of {\sc Use Cases} and {\sc Constraints}. 
A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has its name, description, and which ones should be checked during the execution of the application. 
Each {\sc Constraint} is represented as a stereotyped ({\sf constraint}) use case. 
There are three kinds of constraint:
\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value''; 
\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and 
\textit{(iii)} A restriction may describe {\sc Exceptional Behaviour} constraints, represented as the stereotype {\sf ``exceptional\_behaviour''}.
These concepts are present in the following example.

@@

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}. The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service. The buy music pre-conditions consist in verifying: 
(i) if the User data are correct; 
(ii) if the User is already logged in Spotify; 
(iii) if bank account information are correct and; 
(iv) if there are enough funds in the bank account to cover the payment. 
A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization. There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified. To update services, depending of each service, there may be different restrictions. As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter. In the case of Facebook, it is required that the user is already logged in Spotify and these data are the same as Facebook. 
As post-condition, the application ensures that the Facebook service sends a notification of success. 
To update Twitter a pre-condition is required, while to update Facebook it is necessary to check a pre-condition and a confirmation notice (modelled as post-condition). 
As a pre-condition for ``twitter update'' it is necessary that (i) the music format be correct and (ii) the twitter login and password be correct for the update.
\end{example}

@@

As shown in Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model 
provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
to group contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a security policy.
 The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action. 
An instance of this meta-class is graphically represented as a partition in the activity diagram. 
A collaborator can be either internal or external to the system. 
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

@@

\begin{figure}[t]
\centering 
\includegraphics[width=1.0\textwidth]{figs/PiServiceComposition} 
\caption{$\pi$-Service Composition Metamodel.}
\label{fig:e-scomposition-metamodel}
\end{figure}

@@

\item In order to represent constraint types associated to services compositions, we introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules. 
Intuitively, the conditions of each rule will be evaluated.
In case of no compliance, the actions defined by the rule will be performed.
The {\sc Rule} meta-class represents the types of event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
\end{itemizedTrivlist}
Instances of this meta-model are represented as UML activity diagrams. 
\begin{figure}[t]%[htpb]
\centering
\includegraphics[width=0.95\textwidth]{figs/piServiceComposition-toPublishMusic}
%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
\caption{$\pi$-ServiceComposition Model for the ``To publish
music'' business service.}
\label{fig:servicecompositionmodel}
\end{figure}

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
 We model a business process that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
Both {\em Facebook} and {\em Twitter} require authentication. 
Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
%{\color{red} To be completed!.}
In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the user space. 
%A call to such services must be part of the authentication protocol required by these services.
In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}). 
In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
{\sf\small OAuthPolicy} will implement the open authentication protocol.
The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed. 
Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
On the notification of an event of that type, a credential is obtained using the username and password. 
\hfill\openbox
\end{example}

@@

This level focusses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system. At the PSM level we have lower-level models that can be automatically translated into actual computer programs. 
We have defined one meta-model at this level.

The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions. 
Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.

@@

\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
A brief description of these classes is given next.
\end{itemizedTrivlist}
%
\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Metamodel.}
\label{fig:PPEWSmetamodel}
\end{figure}

>>>>

This section presents \pisodm, an MDD-based methodology for building service compositions with non-functional requirements. 
\pisodm provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}). 
Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
The refinement process is driven by transformation rules specified between the meta-models.

@@

This level focuses on the highest-level view of the system, including its business and requirement specifications.
At this stage of the development, the structure and system processing details are still unknown or undetermined.  
\pisodm uses the \textit{e$^3$value}~\cite{Gordijn02valuebased} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.

@@

\begin{example}[To Publish Music]\label{ex:toPublicMusic}
Let us consider the scenario ``To Publish Music'', used as a running example in this section:
An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts. 
In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
The following (external) services will be used by the application:
\begin{itemize}
\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item Facebook and Twitter services export methods for  updating the status of a given user:
\begin{itemize} 
%\item{\sf\small get-Status ( usedid ): String ; 
\item {\sf\small update-Status ( userid, new-status ): String}; 
\end{itemize}
\end{itemize}

@@

The CIM level models are the basis for the development of the application. 
The information presented at this level is (manually) refined into PIM-level models of the methodology \pisodm described next.
Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.

@@

This level focuses on the system functionality, hiding the details of any particular platform.
The specification defines those parts of the system that do not change from one platform to another. 
Our methodology defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess} and \textit{$\pi$-ServiceComposition}.

@@

Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-functional requirements.
The notion of \textit{policy} is used to describe NFRs. 
(In Figure~\ref{fig:CIM:usecasemetamodel}, we have highlighted the concepts related to NFRs).
The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing non-functional requirements with  the concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc non-functional requirement}, {\sc non-functional attribute} and {\sc Constraint}. An {\sc End Consumer} is represented by an {\sc Actor}. An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
 The {\sc Business Collaborator} concept is represented  through {\sc Packages}. 
A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modeled. 
Each {\sc Business Collaborator} combines the features described in each  {\sc Package}. 
Thus, the services functions can be specified as being grouped into packages.

@@

 The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented as part of description of {\sc Use Cases} and {\sc Constraints}. 
A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has its name, description, and which ones should be checked during the execution of the application. 
Each {\sc Constraint} is represented as a stereotyped ({\sf constraint}) use case. 
There are three kinds of constraint:
\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value''; 
\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and 
\textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
These concepts are present in the following example.

@@

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}. The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service. The buy music pre-conditions consist in verifying: 
(i) if the User data are correct; 
(ii) if the User is already logged in Spotify; 
(iii) if bank account information are correct and; 
(iv) if there are enough funds in the bank account to cover the payment. 
A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization. There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified. To update services, depending of each service, there may be different restrictions. As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter. In the case of Facebook, it is required that the user is already logged in Spotify and these data are the same as Facebook. 
As post-condition, the application ensures that the Facebook service sends a notification of success. 
To update Twitter a pre-condition is required, while to update Facebook it is necessary to check a pre-condition and a confirmation notice (modeled as post-condition). 
As a pre-condition for ``twitter update'' it is necessary that (i) the music format is correct and (ii) the twitter login and password are correct for the update.
\end{example}

@@

As shown in Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model 
provides meta-classes to represent workflows\footnote{Workflows are transformed into implemented service compositions.} that model  business processes.
The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
to group contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a security policy.
 The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action. 
An instance of this meta-class is graphically represented as a partition in the activity diagram. 
A collaborator can be either internal or external to the system. 
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

@@

\begin{figure}[t]
\centering 
\includegraphics[width=1.0\textwidth]{figs/PiServiceComposition} 
\caption{$\pi$-Service Composition Meta-model.}
\label{fig:e-scomposition-metamodel}
\end{figure}

@@

\item In order to represent constraint types associated to services compositions, we introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules. 
The conditions of each rule are evaluated and, if they are not verified, the actions of the rule will be performed.
The {\sc Rule} meta-class represents the types of event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
\end{itemizedTrivlist}
Instances of this meta-model are represented as UML activity diagrams. 
\begin{figure}[t]%[htpb]
\centering
\includegraphics[width=0.95\textwidth]{figs/piServiceComposition-toPublishMusic}
%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
\caption{$\pi$-ServiceComposition Model for the ``To publish
music'' business service.}
\label{fig:servicecompositionmodel}
\end{figure}

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
 We model a business process that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
Both {\em Facebook} and {\em Twitter} require authentication. 
Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
%{\color{red} To be completed!.}
In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that read and update the user data. 
%A call to such services must be part of the authentication protocol required by these services.
In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}). 
In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
{\sf\small OAuthPolicy} implements the open authentication protocol.
The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed. 
Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
On the notification of an event of that type, a credential is obtained using the username and password. 
\hfill\openbox
\end{example}

@@

This level focuses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system. At the PSM level we have lower-level models that can be automatically translated into actual computer programs. 
We have defined one meta-model at this level.

The \textit{$\pi$-PEWS} meta-model provides concepts for modeling service compositions. 
Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.

@@

\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
A brief description of these classes is given next.
\end{itemizedTrivlist}
%
\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Meta-model.}
\label{fig:PPEWSmetamodel}
\end{figure}

### 2014-05-07 JournalToSubmit2013/transformationRules.tex
Final version?
final version (V2)

\pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models. 
These rules are used to build a semi-automatic tool.
The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.

@@

A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
The set of resulting assertions  are grouped into a {\sf Contract}.
Constraints are transformed according to their type:
{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} attribute set to false are transformed into {\sf Assertion}s;
{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behaviour}s.

@@

The ``listen music'' use case is transformed into a Service Action. 
This Action  represents a Spotify service function that can be invoked to play the music. 
For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}. 
The ``download music'' use case  includes the payment process to buy the music. 
Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service Activity} that aggregates these {\sf Actions}.   
This \textsf{Service Activity} is transformed into a sequence flow on the $\pi$-service process model.
%(as shown in figure \ref{fig:transformation-example-include}). 
The same rule is applied for the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
 \end{example}

@@

Some elements of the $\pi$-service composition model are obtained from the $pi$-UseCase model: 
A {\sf Package} in the $\pi$-use case model is transformed into a {\sf Business Collaborator};
\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Funtional requirements} of the $\pi$-ServiceComposition model. 
These requirements will be associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).

@@

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.

The "securityLoginPolicy" consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
The information about Facebook and Spotify (both of them {\sf Business
Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
model.
\end{example}

@@

Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Preconditions}, {\sf Postcondition} and {\sf TimeRestrictions}.

@@

\subsection{Implementation}
We have provided tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.

>>>>

\pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models. 
These rules are used to build a semi-automatic tool (the \pisodm environment plugin for Eclipse).
The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.

@@

A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
The set of resulting assertions  are grouped into a {\sf Contract}.
Constraints are transformed according to their type:
{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} attribute set to false are transformed into {\sf Assertion}s;
{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.

@@

The ``listen music'' use case is transformed into a Service Action. 
This Action  represents a Spotify service function that can be invoked to play the music. 
For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}. 
The ``download music'' use case  includes the payment process to buy the music. 
Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service Activity} that aggregates these {\sf Actions}.   
This \textsf{Service Activity} is transformed into a sequence flow on the $\pi$-service process model.
%(as shown in figure \ref{fig:transformation-example-include}). 
The same rule is applied to the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
 \end{example}

@@

Some elements of the $\pi$-service composition model are obtained from the $\pi$-UseCase model: 
A {\sf Package} in the $\pi$-use case model is transformed into a {\sf Business Collaborator};
\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Functional requirements} of the $\pi$-ServiceComposition model. 
These requirements are associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).

@@

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
The ``securityLoginPolicy'' consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
The information about Facebook and Spotify (both of them {\sf Business
Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
model.
\end{example}

@@

Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Pre-conditions}, {\sf Post-condition} and {\sf TimeRestrictions}.

@@

\subsection{Implementation}
We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
